---
title: "Understanding Error Messages in R - Part 3"
author: "William Sheahan"
date: "4/18/2022"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Example of a Novel Error Message**

This week, we will be taking a crack at a brand-new error message
produced by routine code when trying to determine the number of cases in
each district of Eastern Province in 2021. This time, I will not be
providing an answer in this document. Please use the steps from the last
two trainings to figure out what went wrong and how to fix it.

```{r new error, warning=FALSE, message=FALSE, error = TRUE}
#load packages
library(tidyverse)
library(lubridate)

#set your working directory using your own filepath, not mine
setwd("C:/Users/wsheahan/OneDrive - PATH/Desktop/r-workshop")

#load datasets
case_data <- read_csv("data/district-cases-long.csv")

#generate summary
case_data %>%
  filter(province == "Eastern") %>%
  filter(between(period, as.Date("2021-01-01"), as.Date("2021-12-01"))) %>%
  group_by(district) %>%
  sum(count, na.rm = TRUE)

```

What went wrong? Use the error message above to determine the source of the error, and follow the steps from the previous lesson to fix it. Tips and tricks from previous lessons below. 

## **Tips and Tricks**

This is a quick rundown of my recommended "process" for dealing with
error messages in R. The first four steps are a miniature version of the
lessons we have already covered in this unit. Steps 5 through 7 are more
broad strategies, with commentary from [Y. Wendy Huynh's R for Graduate
Students](https://bookdown.org/yih_huynh/Guide-to-R-Book/trouble.html)
that I find to be particularly helpful.

1.  First, I usually like to look for misspellings or improper calls to
    the wrong data objects.

    -   This tends to account for well over 90% of the errors in my own
        code (Because I copy and paste a lot from one code section to
        another, I am particularly vulnerable to hanging parentheses, +
        signs, and dplyr %>% pipes)

2.  Second, I look at the dataset itself to see if I can find any clues
    as to what went wrong in the dataset structure.

    -   Has a column gotten filled with NA values somehow?

    -   Did a variable get misspelt or removed by mistake?

        -   Try using the *table()* command

    -   Is the variable type incompatible with my code?

        -   Try using the *class()* command

3.  Third, I will run the code line by line until I find exactly where
    the error occurs.

    -   This is particularly useful when using dplyr pipes %>% over
        multiple lines of code

4.  Fourth, I will google the problem.

    -   Use the formula: "key word" "error message" "R"; and look for
        stack overflow responses that match your situation to some
        degree.

    -   No answer is likely to give you exactly the solution to your
        situation, find the piece that is useful to you.

5.  Fifth, I will restart R.

    -   "In rare (but real) circumstances, I have found that RStudio can
        run into some issues that have nothing to do with your code. If
        your code is running strangely, the age-old trick of turning it
        off and on again (i.e., RStudio or your computer) occasionally
        does work. Make sure you save the information you need to save
        before you close RStudio or restart your computer. Upon
        reopening RStudio, make sure the Global Environment is clean
        before attempting to execute code." (Huynh, 2019)

6.  Sixth, I will take a break.

    -   "Another option is to take a break. During that period, little
        elves come and fix your computer for you. If you don't leave,
        they'll never show up. In all seriousness, error messages can be
        anger-inducing and in your cloud of frustration, you may not be
        able to think clearly. Step away for a bit and look at it from a
        different emotional state." (Huynh, 2019)

7.  Seventh, I will ask someone for help.

    -   "Finally, ask someone for help. An experienced R user should be
        able to figure out what went wrong pretty quickly. This is
        because experts have made (and hopefully fixed) 100X more errors
        than beginners; many people struggle with the same problems at
        the beginning. Problem solving is (in my opinion) the best way
        to learn. However, it's also the most time consuming. Ask for
        help when you need it. Time is an invaluable, nonrenewable
        resource." (Huynh, 2019)
