---
title: "Understanding Error Messages in R - Part 1"
author: "William Sheahan"
date: "3/22/2022"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, warning=FALSE,message=FALSE}
library(tidyverse)
library(lubridate)
```

## **Understanding Error Messages**

One of the most difficult and perplexing things about learning R is
learning how to decode error messages. Usually, when something in our
code doesn't work, R tries to be helpful by telling us went wrong, using
an error message. Unfortunately, the meaning of these messages is not
always very clear.

In this unit, we are going to learn about some of the most common R
error messages, why they occur, how to identify their source, and how to
fix them. Over the course of the next few sessions, we will take on more
and more complex errors that can manifest in our code and learn how to
effectively search the internet for answers to our more complex coding
problems.

To begin, let's go over which aspects of R code cause the most common
kinds of errors.

## *Spelling and Punctuation Errors*

**Misspelling** - By far the most common error in my own code, and a
likely issue in just about any situation, is an accidental misspelling
of an object or function that has been overlooked. If you get an error
message saying that something *"does not exist"*, *"cannot be opened"*,
or *"could not find function"*, there is a good chance that a letter has
been misplaced.

Double check the lines of code for any obvious misspellings, **running
each line one at a time** to identify the place where the error occurs,
and then if you cannot immediately identify the error, try to call up a
table of the variables you are working with, and look at the output to
see if there is an unexpected spelling in the dataset itself that you
need to fix or match in your code.

```{r loading data, warning=FALSE, message=FALSE, error = TRUE}
setwd("C:/Users/wsheahan/OneDrive - PATH/Desktop/r-workshop")
case_data <- readr::read_csv("data/district-cases-long.csv")
```

```{r example - misspellings, error = TRUE}
case_data %>% 
  filter(data_type %in% c("Clinical", "Confirmed", "Confirmed_Passive_CHW")) %>%
  group-by(lubridate::year(period), district) %>%
  summarise(count)

```

When we run the above code chunk, we get an error saying that a function
"group" was not found.

*Question 1:* Why did this happen?

*Question 2:* How would we fix it?

**Capitalization** - It is important to remember that R is
*case-sensitive*, meaning that it does matter whether a letter is
**Capitalized** or **lower-case**. This is important for calling
datasets and variable names as well as when typing out a function or a
graphical parameter. If you have an error saying something doesn't exist
or can't be found, and you have checked already that it is spelled
correctly, make sure you didn't accidentally give it a capital or
lower-case beginning by mistake.

```{r example - capitalization, error = TRUE}
case_data %>% 
  filter(data_type %in% c("Clinical", "Confirmed", "Confirmed_Passive_CHW")) %>%
  group_by(lubridate::year(period), District) %>%
  summarise(count)

```

When we run the above code chunk, we get an error with a lot of
information. In this case, the first part of the error message "adding
computed columns in group_by() x Problem with 'mutate()' input '...2'"
is somewhat hard to understand. Let's focus on the last part of the
error message "object 'District' is not found". This tells us that R
searched for an object, in this case a variable we want to group by, and
couldn't find it, even though we know that district is a column we have
in our dataset. Try running a names() command on your dataset to see
what the column names are.

*Question 3:* Why did this happen?

*Question 4:* How do we fix it?

**Closing Punctuation** - It is important to always double-check your
parentheses (), brackets \[\], and commas , in R because every time R
sees an open parentheses or bracket, it will continue to assume
everything after that is part of the same statement until the
parentheses or bracket is closed. This is most often an issue in the
case where we have multiple open brackets and only remember to close
one.

```{r example - closing punctuation, error = TRUE}
case_summary <- case_data %>% 
  filter(data_type %in% c("Clinical", "Confirmed", "Confirmed_Passive_CHW")) %>%
  group_by(lubridate::year(period, district) %>%
  summarise(count)

table(case_summary$district)
```

For some reason, when we try to run this code chunk we get the above
error of "unexpected symbol in: table". This is a very common error, and
it usually indicates a piece of punctuation is missing or there is one
too many punctuation marks.

*Question 5:* Why did this happen?

*Question 6:* How would we fix it?

**Continuing Punctuation** -

```{r example - continuing puntuation, error = TRUE}
case_summary <- case_data %>% 
  filter(data_type %in% c("Clinical", "Confirmed", "Confirmed_Passive_CHW")) %>%
  group_by(lubridate::year(period), district) %>%
  summarise(count) %>%

table(case_summary$district)
```

Note that after running the first chunk of the above code, no error
message is initially produced, but also no summary. If you look in your
console (bottom left of the screen), you will see a continuing "+" that
indicates R thinks the code chunk is still ongoing. If you attempt to
run the entire chunk of code, you get an error message saying that your
"object 'case-summary' not found".

*Question 7:* Why did this happen?

*Question 8:* How would we fix it?

## **Missing Objects**

**Filepath Issues** - One of the earliest problems we encounter when
coding is getting datasets to load properly in our session. We have
tried to take steps to prevent this issue by working within a project,
essentially a dedicated working directory that tells R where files are
stored. If you are working within a project, you should only have to use
a relative filepath (as in the early part of this code where we call
"data/district-cases-long.csv").

If you are not working from within a project, you have to specify the
full filepath for an item, or manually set a working directory to where
most of your files are stored. If any part of this filepath is incorrect
or missing, you will encounter issues accessing your data.

For our example, please try to download the Example_Cleaning_Jan_10.csv
file in the google drive "Example Data" folder here -
<https://drive.google.com/file/d/1fKbTh9rCnG79Q4J4D2gpXC5DjjEZrfAd/view?usp=sharing>

```{r example - filepath issue, error = TRUE}
example_data <- read_csv("C:/Users/wsheahan/OneDrive -
                        PATH/Desktop/r-workshop/data/Example_Cleaning_Jan_10.csv")

```

If you try to run my original code, the filepath will not work for you.
You will likely get an error along the lines of "Error: 'filepath.csv'
does not exist". If you are unsure of the exact filepath for your data,
try writing out everything up to the quotation marks from the above
chunk, then hitting the *tab* key inside the "". This should autofill
with datasets in your working directory so that you can choose the
Example_Cleaning_Jan_10.csv dataset if you stored it in your project's
data folder.

*Question 9:* Why did this happen?

*Question 10:* How would we fix it?

**Unloaded libraries** - Another common error message to receive is that
a function we are trying to use doesn't exist. Usually, "error in
*function*(.) could not find function(*function*)". First, it is
important to check that we have spelled the function name correctly,
including any underscores or other symbols. If the function name is
spelled correctly, that likely means that we haven't loaded the package
that the function comes from using the *library()* command around the
package's name.

```{r example - library issue, error = TRUE}
example_data <- read_csv("C:/Users/wsheahan/OneDrive -
                        PATH/Desktop/r-workshop/data/Example_Cleaning_Jan_10.csv") %>%
  clean_names()

```

If you run the above code chunk with your own correct filepath, you
should get an error saying "could not find function "clean_names", even
though we have used the clean_names() function in a previous session.

*Question 11:* Why did this happen?

*Question 12:* How would we fix it?

## **Summary**

This list of common errors is not comprehensive. The reality of coding
in R is that every keystroke matters to some degree, and that every
letter or number you type has the potential to cause an error. Luckily,
R's error messages usually do a pretty good job of telling us what we
have done wrong, and give us a good starting point for fixing our
mistakes.

I want to leave you with a quote from a blog post I found on
interpreting common R errors.

"Almost all of these types of errors have a common theme: **R is looking
for something that isn't there.** If you look through the errors above,
you'll see R is mostly reporting that it expects to find an object, a
logical value, a function, a file, a sub-object, or a dependency based
on the users' input, but that thing isn't there. This is an important
concept to convey in any "error interpretation" lesson." -
<https://warin.ca/posts/rcourse-howto-interpretcommonerrors/>

A key takeaway from this lesson should hopefully be that the most common
errors are often very small (usually a single letter) discrepancies that
prevent R from making a connection with our dataset or with a loaded
function or package. Understanding this paradigm is important for
tackling problem solving as a systematic process in R (our next lesson),
because it allows us to think from the point of view of R, and to see
where errors are likely to be generated so we can actively prevent them
going forward in our own code.

When it comes to understanding, solving, and preventing error messages
in R there is no substitute for practice and experience. That is why we
are taking the next few sessions to begin the process of working with
error messages in R, so that we all have a strong base-layer of
understanding to use in our own experiences going forward.

## **Tips and Tricks**

1.  When you encounter an error, **read the whole message and look for
    clues**. Usually it will try to tell you where the error occurred
    (Example: if it says "could not find function "clean_names" then you
    know there is probably an error in that line where you call the
    clean_names() function).

2.  If you aren't sure where an error occurred, **run the code line by
    line** until you find it. This works especially well using dplyr's
    pipe system that isolates code steps on different lines.

3.  If you are having filepath issues, try using **tab complete** to
    fill in your filename automatically.

4.  If misspellings are possibly the culprit, try using
    **names(dataset)** to get the spellings of all columns in the
    dataset, or **table(dataset\$variable)** to get the spellings of all
    possible values for a given variable

5.  If you are having lots of small issues, it is usually helpful to
    **keep all of your code visible** by keeping individual lines short
    enough to stay on your visible screen (usually less than 80
    characters). If you find a line is stretching off the screen, insert
    a line break. It's hard to fix what you can't see.
